Effectful Semantics for Substructural SSA

### 2023-11-07 pldi24/paper.tex
Twiddle, bump Typst version
Tap

@@

\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to I\) is a map from streams of \(M\) to \(I\) satisfying
  \(
    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
  \),
  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the stream \(\ell\).
\end{definition}
Some basic examples include:
\begin{enumerate}
  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts on \(\nats \cup \{\infty\}\) by addition.
\end{enumerate}
We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\). For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq \omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) = \{0\} \subseteq Z(\nats) = \nats\). Generally, \(\mc{SZ}(M, I)\) always contains the unit of \(M\).

\begin{definition}[Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Pi\;A = A \times M + I\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and \(\mu_A = [\lambda (a, m). m \cdot a, \ms{id}]\)
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\).
\end{definition}

@@

\begin{definition}[Nondeterministic Trace Monad]
  Given a stream action \((\Pi, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\{\iota_0 (a, 1)\}\)
  and \(\mu_A = ...\)
  This is an Elgot monad, with
  \[f^\dagger(a) = \{\} \cup \{\} \cup \{\}\]
\end{definition}

@@

Given a set \(A\), we can define a predicate \(\varphi\) on \(A\), \(\mb{Heap}\) pairs to have the following type:
\TODO{this}
\TODO{Objects \((A, \varphi)\), \((B, \psi)\)}
\TODO{arrows}
\TODO{note on magic wand}
\TODO{Tensor product and separating conjunction}
\TODO{Coproduct, Elgot structure inherited from base category}
\TODO{sliding w/ properly restricted equality}

@@

\TODO{equality vs \cite{mellies-ftrs}}
\TODO{potential solutions}

>>>>

We will begin by giving a useful foundation for a wide class of \isotopessa models: the "trace models", which accumulate a (potentially infinite) nondeterministic trace of monoidal "events". This will later come in handy for reasoning about output behaviour and weak memory.

We begin with some definitions:
\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to I\) is a map from streams of \(M\) to \(I\) satisfying
  \(
    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
  \),
  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the stream \(\ell\).
\end{definition}
Some basic examples include:
\begin{enumerate}
  \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\) forms a stream action on the one element set \(\mathbf{1}\).
  \item \todo{product, giving monoidal structure?}
  \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\), where \(\ell \cdot s\) prepends the (finite) list \(\ell\) to the (potentially infinite) sequence \(s\).
  \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts on \(\nats \cup \{\infty\}\) by addition.
\end{enumerate}
We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\). For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq \omega}) = Z(A^*) = \{[\,]\}\), while \(\mc{SZ}(\nats, \nats \cup \{\infty\}) = \{0\} \subseteq Z(\nats) = \nats\). Generally, \(\mc{SZ}(M, I)\) always contains the unit of \(M\).

We can now define the \textit{trace monad} of a stream action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A \times M + I)\), with
  \(\eta_A = \lambda a.\iota_0 (a, 1)\)
  and 
  \[\mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}\]
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') = \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). This is an Elgot monad, with iteration operator
  \(f^\dagger(a) = \bigcup_{n \in \nats \cup \{\infty\}}f_n(a)\), where
  \TODO{this}

  \TODO{deterministic trace monad; inherited Elgot structure: use as example of "Elgot submonad"? Go cite \cite{carette-central-2022}}
\end{definition}

@@

\TODO{merge det/nondet}

@@

@@

Given a set \(A\), we can define a predicate \(\varphi\) on \(A\), \(\mb{Heap}\) pairs to have the following type:
\TODO{this}
\TODO{Objects \((A, \varphi)\), \((B, \psi)\)}
\TODO{arrows}
\TODO{note on magic wand}
\TODO{Tensor product and separating conjunction}
\TODO{Coproduct, Elgot structure inherited from base category}
\TODO{sliding w/ properly restricted equality}
\TODO{equality vs \cite{mellies-ftrs}}
\TODO{potential solutions}

@@
